\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2012-08-14 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
using knitr instead of pgfSweave and misc vignette updated
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@68442 bc3139a8-67e5-0310-9ffc-ced21a209358

\SweaveOpts{echo=T,eval=T,fig=F,results=verbatim,cache=F,tikz=T,external=F,prefix.string=sweave-cache/figs/fig}

@@

\hypersetup{%
  hyperindex = {true},
  colorlinks = {true},
  linktocpage = {true},
  plainpages = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  pdfstartview = {Fit},
  pdfpagemode = {UseOutlines},
  pdfview = {XYZ null null null}
}

@@

%% <<setup,cache=FALSE>>=
%%  setCacheDir("sweave-cache/values")
%% @

<<environment,cache=FALSE,echo=FALSE>>=
library("MSnbase")
library("grid")
library("reshape2")
@

@@

\begin{figure}[!ht]
<<spectrum-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=5>>=
plot(sp, reporters = iTRAQ4, full = TRUE) 
@ 
\caption{Raw MS2 spectrum with details about reporter ions.}
\label{fig:spectrum-plot}
\end{figure}

<<bsasel,echo=FALSE>>=
bsasel <- fData(itraqdata)$ProteinAccession == "BSA"
@

@@

\begin{figure}[p]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(bsa, reporters = iTRAQ4, full = FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. 
  See section \ref{sec:norm} to rescale peaks identically. }
\label{fig:msnexp-plot}
\end{figure}

@@

\begin{figure}[htb]
  \begin{center}
    \includegraphics[width=\linewidth]{./plot2d-figure.png}
\caption{ Illustration of the \Rfunction{plot2d} output. } 
\label{fig:plot2d}
  \end{center}
\end{figure}

@@

\begin{figure}[htb]
  \begin{center}
    \includegraphics[width=\linewidth]{./plotDensity-figure.png}
\caption{ Illustration of the \Rfunction{plotDensity} output. } 
\label{fig:plotDensity}
  \end{center}
\end{figure}

@@

There are several methods implemented to perform basic data manipulation. 
Low intensity peaks can be set to 0 with the \Rfunction{removePeaks} method 
from spectra or whole experiments. The intensity threshold below which peaks are removed 
is defined by the \Robject{t} parameter. \Robject{t} can be specified directly 
as a numeric. The default value is the character \texttt{"min"}, that will remove 
all peaks equal to the lowest non null intensity in any spectrum. 
We observe the effect of the \Rfunction{removePeaks} method by comparing 
total ion count (i.e. the total intensity in a spectrum) with the \Rfunction{ionCount} 
method before (object \Robject{itraqdata}) and after (object \Robject{experiment}) for 
spectrum \texttt{X55}.

@@

<<prepare-spectrum-plot2,echo=FALSE>>=
p1 <- plot(itraqdata[["X55"]], full = TRUE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]], full = TRUE) + theme_gray(5) 
@

\begin{figure}[!ht]
<<spectrum-plot2,fig=T,echo=F,tikz=T,external=F,width=5,height=2.5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p1,vp=vplayout(1,1))
print(p2,vp=vplayout(1,2))
@ 
\caption{Same spectrum before (left) and after setting peaks <= 400 to 0.}
\label{fig:spectrum-plot2}
\end{figure}

@@

<<clean,echo=TRUE,cache=FALSE>>=
## number of peaks
peaksCount(itraqdata[["X55"]])
peaksCount(experiment[["X55"]])
experiment <- clean(experiment, verbose = FALSE)
peaksCount(experiment[["X55"]])
@

<<preprosp,cache=FALSE,echo=FALSE>>=
int <- c(0,1,1,3,1,1,0,0,0,1,3,7,3,1,0)
mz <- c(113.9,114.0,114.05,114.1,114.15,114.2,114.25,
        114.3,114.35,114.4,114.42,114.48,114.5,114.55,114.6)
ppsp <- new("Spectrum2",intensity=int,mz=mz,centroided=FALSE)
p1 <- plot(ppsp,full=TRUE) + theme_gray(5) + 
  geom_point(size=3,alpha=I(1/3)) +
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Original spectrum") 
p2 <- plot(removePeaks(ppsp,t=3),full=TRUE) + 
  theme_gray(5) +
  geom_point(size=3,alpha=I(1/3)) + 
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Peaks < 3 removed") 
p3 <- plot(clean(removePeaks(ppsp,t=3)),full=TRUE) + 
  theme_gray(5) +
  geom_point(size=3,alpha=I(1/3)) + 
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Peaks < 3 removed and cleaned") 
@

\begin{figure}[p]
  \begin{center}
<<preproc-plot,fig=T,echo=F,tikz=T,external=F,width=3,height=6>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectively deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
  \end{center}
\end{figure}

@@

<<simplesp,cache=FALSE,echo=FALSE>>=
int <- c(0,1,1,3,1,1,0)
mz <- c(113.9,114.0,114.05,114.1,114.15,114.2,114.25)
ssp <- new("Spectrum2",intensity=int,mz=mz,centroided=FALSE)
p <- plot(ssp,full=TRUE)
p <- p + theme_gray(5)
@

\begin{figure}[p]
<<quantitation-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p + opts(title = "Quantitation using 'sum'") +
      geom_point(size = 3, alpha = I(1/3), colour = "red"),
      vp = vplayout(1, 1))
print(p + opts(title = "Quantitation using 'max'") +
      geom_point(aes(x = 114.1, y = 3),
                 alpha = I(1/18),
                 colour = "red", size = 3),
      vp = vplayout(1, 2))
print(p +
      opts(title = "Trapezoidation and strict=FALSE") +
      geom_polygon(alpha = I(1/5), fill = "red"),
      vp = vplayout(2, 1))
print(p +
      opts(title = "Trapezoidation and strict=TRUE") + 
      geom_polygon(aes(x = c(NA, 114.05, 114.05, 114.1, 114.15, 114.15, NA),
                       y=c(NA, 0, 1, 3, 1, 0, NA)), fill = "red", alpha = I(1/5)),
      vp = vplayout(2,2))

@@

\begin{figure}[p]
<<norm-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=7>>=
.plot <- function(x,ttl=NULL) {
  boxplot(exprs(x),
          main=ifelse(is.null(ttl),processingData(x)@processing[2],ttl),
          cex.main=.8,
          cex.lab=.5,
          cex.axis=.5,
          cex=.8)
  grid()
}
oldmar <- par()$mar
par(mfrow=c(3,2),mar=c(2.9,2.9,2.9,1))
.plot(qnt,ttl="Non-normalised data")
.plot(qnt.max)
.plot(qnt.sum)
.plot(qnt.quant)
.plot(qnt.qrob)
.plot(qnt.vsn)
@ 
\caption{ Comparison of the normalisation \Robject{MSnSet} methods. 
  Note that vsn also glog-transforms the intensities. }
\label{fig:norm-plot}
\end{figure}

@@

\begin{center}
\begin{figure}[!ht]
<<cv-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=4>>=
print(p)
@ 
\caption{CV versus signal intensity comparison for log2 and vsn
  transformed data. Lines indicate running CV medians.}
\label{fig:cv-plot}
\end{figure}
\end{center}

@@

<<prepare-msnset-nrom-plot,cache=FALSE,echo=FALSE>>=
p <- plot(normalise(experiment[bsasel], "max"),
          reporters = iTRAQ4, full = FALSE)
p <- p + theme_gray(7)
@

\begin{figure}[p]
<<msnexp-norm-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=6>>=
print(p)
@ 
\caption{Experiment-wide normalised MS2 spectra. The y-axes of the individual 
  spectra is now rescaled between 0 and 1 (highest peak), as opposed to 
  figure \ref{fig:msnexp-plot}. }
\label{fig:msnexp-norm-plot}
\end{figure}

@@

\begin{figure}[!hbt]
  \begin{center}
<<incompdiss-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=2.5>>=
dfr <- melt(exprs(incompdiss))
colnames(dfr) <- c("feature", "reporters", "intensity")
dfr$reporters <- sub("iTRAQ5.", "", dfr$reporters)
repsum <- apply(exprs(incompdiss), 1, function(x) sum(x[1:4]))
dfr2 <- data.frame(iTRAQ1to4 = repsum, iTRAQ5 = exprs(incompdiss)[,5])
p1 <- ggplot(data = dfr, aes(x = reporters,y = log10(intensity))) +
  geom_boxplot() +
  theme_gray(6)
p2 <- ggplot(data = dfr2, aes(x = log10(iTRAQ1to4), y = log10(iTRAQ5))) +
  geom_point(alpha = I(1/2)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  stat_smooth(method = lm, se = FALSE) +
  xlab(label = expression(log[10]~sum~114~to~117)) +
  ylab(label = expression(log[10]~145)) +
  theme_gray(6)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(p1, vp = vplayout(1, 1))
print(p2, vp = vplayout(1, 2))
@ 
\caption{ Boxplot and scatterplot comparing intensities of the 4 reporter ions 
  (or their sum, on the right) and the incomplete dissociation specific peak. }
\label{fig:incompdiss}
  \end{center}
\end{figure}

@@

\clearpage

\section{Session information}\label{sec:sessionInfo} 
<<label=sessioninfo,results=tex,echo=FALSE,cache=FALSE>>=
toLatex(sessionInfo())
@

@@

% LocalWords:  renameSamples

>>>>

@@

\usepackage{geometry}
\geometry{verbose,
  tmargin = 2.5cm,
  bmargin = 2.5cm,
  lmargin = 3.5cm,
  rmargin = 3.5cm}

\usepackage{hyperref}
\usepackage{breakurl}
\hypersetup{%
  pdfusetitle,
  bookmarks = {true},
  bookmarksnumbered = {true},
  bookmarksopen = {true},
  bookmarksopenlevel = 2,
  unicode = {true},
  breaklinks = {false},
  hyperindex = {true},
  colorlinks = {true},
  linktocpage = {true},
  plainpages = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  pdfstartview = {Fit},
  pdfpagemode = {UseOutlines},
  pdfview = {XYZ null null null}
}

@@

<<environment, cache=FALSE, echo=FALSE>>=
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("zoo"))
suppressPackageStartupMessages(library("doMC"))
suppressPackageStartupMessages(library("multicore"))
library("grid")
library("reshape2")
@

<<knitr, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', 
               par = TRUE,
               prompt = TRUE,
               comment = NA)
options(replace.assign = TRUE, 
        width = 68)
knit_hooks$set(par = function(before, options, envir) {
  if (before && options$fig.show != 'none') 
    par(mar = c(4,4,.1,.1),
        cex.lab = .95,
        cex.axis = .9,
        mgp = c(2,.7,0),
        tcl = -.3)
})
@

@@

\begin{figure}[!ht]
<<spectrum-plot, dev='tikz', fig.width=4, fig.height=4, fig.keep='high'>>=
plot(sp, reporters = iTRAQ4, full = TRUE) 
@ 
\caption{Raw MS2 spectrum with details about reporter ions.}
\label{fig:spectrum-plot}
\end{figure}

<<bsasel, echo=FALSE>>=
bsasel <- fData(itraqdata)$ProteinAccession == "BSA"
@

@@

\begin{figure}[p]
<<msnexp-plot, dev='tikz', fig.width=3.7, fig.height=4.5, fig.keep='last'>>=
plot(bsa, reporters = iTRAQ4, full = FALSE) + theme_gray(8)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. 
  See section \ref{sec:norm} to rescale peaks identically. }
\label{fig:msnexp-plot}
\end{figure}

@@

\begin{figure}[htb]
    \includegraphics[width=\linewidth]{./plot2d-figure.png}
\caption{ Illustration of the \Rfunction{plot2d} output. } 
\label{fig:plot2d}
\end{figure}

@@

\begin{figure}[htb]
    \includegraphics[width=\linewidth]{./plotDensity-figure.png}
\caption{ Illustration of the \Rfunction{plotDensity} output. } 
\label{fig:plotDensity}
\end{figure}

@@

There are several methods implemented to perform basic data manipulation. 
Low intensity peaks can be set to 0 with the \Rfunction{removePeaks} method 
from spectra or whole experiments. The intensity threshold below which peaks are removed 
is defined by the \Robject{t} parameter. \Robject{t} can be specified directly 
as a numeric. The default value is the character \texttt{"min"}, that will remove 
all peaks equal to the lowest non null intensity in any spectrum. 
We observe the effect of the \Rfunction{removePeaks} method by comparing 
total ion count (i.e. the total intensity in a spectrum) with the \Rfunction{ionCount} 
method before (object \Robject{itraqdata}) and after (object \Robject{experiment}) for 
spectrum \texttt{X55}. The respective spectra are shown on 
figure \ref{fig:spectrum-clean-plot} (page \pageref{fig:spectrum-clean-plot}).

@@

\begin{figure}[!ht]
<<spectrum-clean-plot, dev='tikz', fig.width=6, fig.height=3, echo=FALSE, fig.keep='high'>>=
p1 <- plot(itraqdata[["X55"]], full = TRUE, plot = FALSE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]], full = TRUE, plot = FALSE) + theme_gray(5) 
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p1,vp=vplayout(1,1))
print(p2,vp=vplayout(1,2))
@ 
\caption{Same spectrum before (left) and after setting peaks <= 400 to 0.}
\label{fig:spectrum-clean-plot}
\end{figure}

@@

<<clean, echo=TRUE, cache=FALSE>>=
## number of peaks
peaksCount(itraqdata[["X55"]])
peaksCount(experiment[["X55"]])
experiment <- clean(experiment, verbose = FALSE)
peaksCount(experiment[["X55"]])
@

<<preprosp, cache=FALSE, echo=FALSE>>=
int <- c(0,1,1,3,1,1,0,0,0,1,3,7,3,1,0)
mz <- c(113.9,114.0,114.05,114.1,114.15,114.2,114.25,
        114.3,114.35,114.4,114.42,114.48,114.5,114.55,114.6)
ppsp <- new("Spectrum2",intensity=int,mz=mz,centroided=FALSE)
p1 <- plot(ppsp, full = TRUE, plot = FALSE) + theme_gray(5) + 
  geom_point(size=3,alpha=I(1/3)) +
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Original spectrum") 
p2 <- plot(removePeaks(ppsp,t=3), full=TRUE, plot = FALSE) + 
  theme_gray(5) +
  geom_point(size=3,alpha=I(1/3)) + 
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Peaks < 3 removed") 
p3 <- plot(clean(removePeaks(ppsp,t=3)), full = TRUE, plot = FALSE) + 
  theme_gray(5) +
  geom_point(size=3,alpha=I(1/3)) + 
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Peaks < 3 removed and cleaned") 
@

\begin{figure}[p]
<<preproc-plot, dev='tikz', fig.width=3, fig.height=6, echo=FALSE>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectively deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
\end{figure}

@@

<<simplesp, cache=FALSE, echo=FALSE, fig.keep='none'>>=
int <- c(0,1,1,3,1,1,0)
mz <- c(113.9,114.0,114.05,114.1,114.15,114.2,114.25)
ssp <- new("Spectrum2",intensity=int,mz=mz,centroided=FALSE)
p <- plot(ssp, full=TRUE, plot=FALSE)
p <- p + theme_gray(5)
@

\begin{figure}[p]
<<quantitation-plot, dev='tikz',fig.width=5, fig.height=5, echo=FALSE>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p + opts(title = "Quantitation using 'sum'") +
      geom_point(size = 3, alpha = I(1/3), colour = "red"),
      vp = vplayout(1, 1))
print(p + opts(title = "Quantitation using 'max'") +
      geom_point(aes(x = 114.1, y = 3),
                 alpha = I(1/18),
                 colour = "red", size = 3),
      vp = vplayout(1, 2))
print(p +
      opts(title = "Trapezoidation and strict=FALSE") +
      geom_polygon(alpha = I(1/5), fill = "red"),
      vp = vplayout(2, 1))
print(p +
      opts(title = "Trapezoidation and strict=TRUE") + 
      geom_polygon(aes(x = c(NA, 114.05, 114.05, 114.1, 114.15, 114.15, NA),
                       y=c(NA, 0, 1, 3, 1, 0, NA)), fill = "red", alpha = I(1/5)),
      vp = vplayout(2,2))

@@

\begin{figure}[p]
<<norm-plot, dev='tikz', fig.width=5, fig.height=7, echo=FALSE>>=
.plot <- function(x,ttl=NULL) {
  boxplot(exprs(x),
          main=ifelse(is.null(ttl),processingData(x)@processing[2],ttl),
          cex.main=.8,
          cex.lab=.5,
          cex.axis=.5,
          cex=.8)
  grid()
}
oldmar <- par()$mar
par(mfrow=c(3,2),mar=c(2.9,2.9,2.9,1))
.plot(qnt, ttl = "Non-normalised data")
.plot(qnt.max, ttl = "Maximum")
.plot(qnt.sum, ttl = "Sum")
.plot(qnt.quant, ttl = "Quantile")
.plot(qnt.qrob, ttl = "Robust quantile")
.plot(qnt.vsn, ttl = "vsn")
@ 
\caption{ Comparison of the normalisation \Robject{MSnSet} methods. 
  Note that vsn also glog-transforms the intensities. }
\label{fig:norm-plot}
\end{figure}

@@

\begin{figure}[!ht]
<<cv-plot, dev='tikz', fig.width=5, fig.height=4, echo=FALSE>>=
print(p)
@ 
\caption{CV versus signal intensity comparison for log2 and vsn
  transformed data. Lines indicate running CV medians.}
\label{fig:cv-plot}
\end{figure}

@@

<<prepare-msnset-nrom-plot, cache=FALSE, echo=FALSE, keep.fig='none'>>=
p <- plot(normalise(experiment[bsasel], "max"),
          reporters = iTRAQ4, full = FALSE, plot = FALSE)
p <- p + theme_gray(7)
@

\begin{figure}[p]
<<msnexp-norm-plot, dev='tikz', fig.width=5, fig.height=6, echo=FALSE>>=
print(p)
@ 
\caption{Experiment-wide normalised MS2 spectra. The y-axes of the individual 
  spectra is now rescaled between 0 and 1 (highest peak), as opposed to 
  figure \ref{fig:msnexp-plot}. }
\label{fig:msnexp-norm-plot}
\end{figure}

@@

\begin{figure}[!hbt]
<<incompdiss-plot, dev='tikz', fig.width=5, fig.height=2.5, echo=FALSE, warning=FALSE>>=
dfr <- melt(exprs(incompdiss))
colnames(dfr) <- c("feature", "reporters", "intensity")
dfr$reporters <- sub("iTRAQ5.", "", dfr$reporters)
repsum <- apply(exprs(incompdiss), 1, function(x) sum(x[1:4]))
dfr2 <- data.frame(iTRAQ1to4 = repsum, iTRAQ5 = exprs(incompdiss)[,5])
p1 <- ggplot(data = dfr, aes(x = reporters,y = log10(intensity))) +
  geom_boxplot() +
  theme_gray(6)
p2 <- ggplot(data = dfr2, aes(x = log10(iTRAQ1to4), y = log10(iTRAQ5))) +
  geom_point(alpha = I(1/2)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  stat_smooth(method = lm, se = FALSE) +
  xlab(label = expression(log[10]~sum~114~to~117)) +
  ylab(label = expression(log[10]~145)) +
  theme_gray(6)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(p1, vp = vplayout(1, 1))
print(p2, vp = vplayout(1, 2))
@ 
\caption{Boxplot and scatterplot comparing intensities of the 4 reporter ions 
  (or their sum, on the right) and the incomplete dissociation specific peak. }
\label{fig:incompdiss}
\end{figure}

@@

\section{Session information}\label{sec:sessionInfo} 
<<label='sessioninfo', results=tex, echo=FALSE, cache=FALSE>>=
toLatex(sessionInfo())
@

@@
